If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-11-03 GroupPolarizationStatmod/paper/method.tex

Our method combines several formal and computational tools to identify which
replications should be retracted and evaluate experimental designs across our
corpus. The first step in the analysis is to find a reasonable pair of latent
null-polarization distributions that appear polarized due to clipping.

\subsection{Identify replications for removal}

The first step is to find which replications of the 57 must be retracted 
due to non-identifiability. We establish non-identifiability by finding one
set of latent null-polarization opinion distributions where the mean doesn't change,
but appear polarized when measured on the ordinal scale used in the experiment.

@@

We formalize this experimental design in the $N\to\infty$ case where the
distribution of ordinal measurements is created by integrating continuous latent
opinion distributions.  Latent opinions in the model are normally distributed with
mean $\mu_t$ and variance $\sigma_t$ for time $t$. We are concerned with only two
times, \emph{pre-} and \emph{post-} deliberation ($t=0$ and $t=T$, respectively).
For short we can write $\mu_\mathrm{pre}$ and $\sigma_\mathrm{pre}$ for
pre-deliberation mean and variance (and similar for \emph{post}-deliberation).  
The latent opinion of participant $i$ at time $t$ is drawn from a normal 
distribution in the model, 
\begin{equation} 
  \omega_{i,t} \sim p(\omega; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution} 
\end{equation} 
\noindent 
Latent opinions are assumed to be normally distributed since they are unbounded 
aggregations of neurobiological activity.

@@

\subsection{Metascience of the false detection rate in group polarization}

Metascience provides a framework for
calculating the false discovery rate across experimental conditions,
aggregated to the level of journal articles and the ten-paper corpus. 
The false discovery rate is then the Type I error rate relative to the rate of
positive discoveries, true or false.
We calculate via simulation and data fitting
the Type I error rate for the experimental designs used in experimental conditions
that, under a preliminary large-$N$ model analysis, generated group
polarization upon ordinal measurement of simple consensus latent opinion 
distributions. These can then be aggregated to estimate the false discovery rate
for each journal article and for the ten-article corpus, which is representative of the
false discovery rate for group polarization studies.

>>>>

Our method combines several formal and computational tools to identify which
replications should be retracted and evaluate experimental designs across our
corpus. The first step in the analysis is to find a reasonable pair of latent
null-polarization distributions that appear polarized due to clipping
(Section~\ref{subsec:identify}). To do this, we develop a formal model and a numeric
rootfinding algorithm to power a Shiny web app we used to interactively identify and
record null-polarization parameters generating such measurement artifacts.

The second step in the analysis is to evaluate the quality of the experimental designs
when we properly account for the ordinal measurement procedure
(Section~\ref{subsec:evaluate}). We use the null-polarization parameters we found in
the first step to generate simulated experimental data, then estimate the difference
between the pre- and post-discussion means using ordered probit models to fit each
distribution—ordered probit models account for the ordinal measurement procedure.  We
estimate the false positive rate and false detection rate for low, medium, and high
significance thresholds in terms of Cohen's $d$. Finally, we use the effect sizes thus
calculated to determine the significance threshold one must use to achieve a 5\% false
positive rate.

\subsection{Identify replications for removal}
\label{subsec:identify}

The first step is to find which replications of the 57 must be retracted due to
non-identifiability. We establish non-identifiability by finding one set of latent
null-polarization opinion distributions where the mean doesn't change, but appear
polarized when measured on the ordinal scale used in the experiment.  In other words,
if we find a null-polarization model that explains the outcome as well as a group
polarization one, the group polarization hypothesis is \emph{non-identifiable} and the
replication must be rejected—the data cannot confirm or disconfirm the target
hypothesis.

@@

% We formalize this experimental design in the $N\to\infty$ case where the
% distribution of ordinal measurements is created by integrating continuous latent
% opinion distributions.

Latent opinions are assumed to be normally distributed in general since they are
aggregations of many neurobiological potentials of
different polarities and magnitudes. We have two opinion distributions for 
two different times in the group polarization experiment process: pre-discussion,
$\tpre$, and post-discussion, $\tpost$.
The 
mean $\mu_t$ and variance $\sigma_t$ for time $t$. We are concerned with only two
times, \emph{pre-} and \emph{post-} deliberation ($t=0$ and $t=T$, respectively).
For short we can write $\mu_\mathrm{pre}$ and $\sigma_\mathrm{pre}$ for
pre-deliberation mean and variance (and similar for \emph{post}-deliberation).  
The latent opinion of participant $i$ at time $t$ is drawn from a normal 
distribution in the model, 
\begin{equation} 
  \omega_{i,t} \sim p(\omega; \mu_t, \sigma_t) = \mathcal{N}(\mu_t, \sigma_t).
  \label{eq:opinionDistribution} 
\end{equation} 
\noindent

@@

We calculate via simulation and data fitting
the Type I error rate for the experimental designs used in experimental conditions
that, under a preliminary large-$N$ model analysis, generated group
polarization upon ordinal measurement of simple consensus latent opinion 
distributions. These can then be aggregated to estimate the false discovery rate
for each journal article and for the ten-article corpus, which is representative of the
false discovery rate for group polarization studies.
